Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update figures

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: |
      | Methodology and Statistics
      | Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Muñoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
  - name: Gerko Vink
    affiliation: |
      | Methodology and Statistics
      | Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(miceheckman)  # for imputation cf heckman models
library(metamisc)     # for ???
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

The true effect is:

@@

```{r echo=FALSE, fig.cap = "True effect in the popularity data"}
fits <-
  purrr::map_dfr(
    unique(popcomp$school),
    ~ {
      lm(
        mod,
        data = filter(popcomp, school == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(school = as.character(.x), .)
  ) %>% filter(term %in% "sex")

n <- popcomp %>% group_by(school) %>% 
  tally() %>% 
  mutate(school = as.character(school))

fits %>% left_join(n, by = "school") %>% 
  ggplot(aes(y = school, x = estimate, linewidth = n)) +
  geom_vline(xintercept = 1, linewidth = 1.5, color = "grey95") +
  geom_linerange(aes(y = school, xmin = conf.low, xmax = conf.high, linewidth = 0.5), position = position_dodge(width = 0.5)) +
  geom_point(shape = 20, position = position_dodge(width = 0.5)) +
  labs(y = "Study", x = "", color = NULL) +
  theme_classic() +
  scale_size(guide = 'none')
```

@@

```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
plot_pattern(popmis, square = TRUE)
```

@@

>>>>

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: |
      | Methodology and Statistics
      | Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Muñoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
  - name: Gerko Vink
    affiliation: |
      | Methodology and Statistics
      | Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, 
      | Amsterdam, The Netherlands

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Sporadic missingness in multilevel data", echo=FALSE, fig.margin=TRUE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(miceheckman)  # for imputation cf. heckman models
library(metamisc)     # for case study data
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

@@

@@

```{r pop_pat, fig.cap = "Missing data pattern in the popularity data", fig.margin=TRUE}
plot_pattern(popmis)
```

@@

# Appendix

### 2023-05-24 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update figures

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(miceheckman)  # for imputation cf heckman models
R> library(metamisc)     # for ???
\end{CodeInput}
\end{CodeChunk}

@@

The true effect is:

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-3-1}

\caption[True effect in the popularity data]{True effect in the popularity data}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

@@

>>>>

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\
Data Analytics and Methods Task Force,\\
European Medicines Agency,\\
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(miceheckman)  # for imputation cf. heckman models
R> library(metamisc)     # for case study data
\end{CodeInput}
\end{CodeChunk}

@@

@@

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}
